---
title: "732A62 Lab 2"
author: "Emil K Svensson & Rasmus Holm"
date: "`r Sys.Date()`"
output:
    pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment 1

# a)

```{r}
library(astsa)
library(kernlab)
library(TSA)
library(forecast)

set.seed(12345)
AR3 <- arima.sim(1000, model = list(order = c(3,0,0),
                                    ar = c(0.8, -0.2, 0.1)))

## The theoretical
AR3.pacf <- pacf(AR3, plot=F)
AR3.data <- ts.intersect(xt = AR3, x1 = lag(AR3, 1), x2 =lag(AR3, 2), x3= lag(AR3, 3))

AR.lm <- resid(lm(xt ~ x1 + x2, data = AR3.data))
AR.lm.lag3 <- resid(lm(x3 ~ x1 + x2 , data = AR3.data))

AR3.pacf[3]

cat(paste("The theoretical value:",round(cor(AR.lm, AR.lm.lag3), digits = 3)))
```
As seen above the theoretical and the output from the pacf-function are very similar. 

# b)

```{r}
set.seed(12345)
AR2 <- arima.sim(100, model = list(order = c(2,0,0),
                                   ar = c(0.8, 0.1)))

ar2.yw <- ar(AR2, order.max = 2, method = "yw", aic = FALSE)
ar2.ols <- ar(AR2, order.max = 2, method = "ols", aic = FALSE)
ar2.mle <- arima(AR2, order = c(2,0,0), method = "ML")

ar2.yw$ar
ar2.ols$ar
ar2.mle$coef

```

The Yule Walker estimate seems to have the parameters closes to the true parameters given in the assignment.

```{r}
ar2.mle 
```


Yes, the theoretical value for $$\phi_2$$ is inside the confidence-intervall for the ML estimate which can seen from the s.e times 1.96 which obviously will cover the true coefficients.

# c)

```{r}
set.seed(12345)
ma.coef <- c(0.3, rep(0, 10), 0.6)
ts4 <- arima.sim(n=200, model=list(order=c(0, 0, 12), ma = ma.coef))

theoretical.acf <- ARMAacf(ma=c(ma.coef, 0.3 * 0.6))
theoretical.pacf <- ARMAacf(ma=c(ma.coef, 0.3 * 0.6), pacf=TRUE)

plot(theoretical.acf, type="h", main="Theoretical ACF")
abline(h=0)

plot(theoretical.pacf, type="h", main="Theoretical PACF")
abline(h=0)

empirical.acf <- acf(ts4)
empirical.pacf <- pacf(ts4)
```

The patterns seem to be somewhat similar. In the theoretical ACF we can see a large spike at lag 14 which also can be seen in the sample ACF. The difference between the two beeing that the sample ACF has some correlation along the lags although under the confidence interval. 

For the PACF we can do the same observation as for the relationship between the theoretical and the sample.




# d)

```{r}
set.seed(12345)
ma.coef <- c(0.3, rep(0, 10), 0.6)
ts5 <- arima.sim(n=200, model=list(order=c(0, 0, 12), ma = ma.coef))

ts5.fit <- arima(ts5, order=c(0, 0, 1), seasonal=list(order=c(0, 0, 1), period=12))
ts5.pred <- predict(ts5.fit, n.ahead=30, se.fit=TRUE)

plot(ts(c(ts5, ts5.pred$pred)), ylim=c(-4, 4))
lines(200 + 1:length(ts5.pred$pred), ts5.pred$pred + 1.96 * ts5.pred$se, lty=2, col="red")
lines(200 + 1:length(ts5.pred$pred), ts5.pred$pred - 1.96 * ts5.pred$se, lty=2, col="red")

gausspr.data <- data.frame(y=ts5, x=1:200)
gausspr.fit <- gausspr(y ~ x, gausspr.data)
gausspr.pred <- predict(gausspr.fit, data.frame(x=201:230))

plot(ts5, xlim=c(0, 230))
lines(c(fitted(gausspr.fit), gausspr.pred), , col="red")
```

In the first plot we can see the MA models predicions seem resonable but after a while the predictions just tails of in to a mean function at y = 0. 

For the gaussian process there is an initial jump in the predictions and afterwards continues with a smoothed pattern similar to the fitted line in the observed data.

# e)

```{r}
set.seed(12345)
ts6 <- arima.sim(model=list(ma=c(0.5), ar=c(0.7)), n=50)

train <- ts(ts6[1:40])
test <- ts(ts6[41:50])

ts6.fit <- arima(train, order=c(1, 0, 1), include.mean = F)
ts6.pred <- predict(ts6.fit, n.ahead=10)

plot(ts(c(train, test)), ylim=c(-4, 7), type="l")
lines(40 + 1:length(test), ts6.pred$pred, col="blue")
lines(40 + 1:length(test), ts6.pred$pred + 1.96 * ts6.pred$se, lty=2, col="red")
lines(40 + 1:length(test), ts6.pred$pred - 1.96 * ts6.pred$se, lty=2, col="red")

plot(40 + 1:length(test), test, ylim=c(-4, 7), xlim=c(40, 50), type="p")
points(40 + 1:length(test), ts6.pred$pred, col="blue")
lines(40 + 1:length(test), ts6.pred$pred + 1.96 * ts6.pred$se, lty=2, col="red")
lines(40 + 1:length(test), ts6.pred$pred - 1.96 * ts6.pred$se, lty=2, col="red")
```

The first of the 10 observations withheld from the fitted model is the only observation that is not in the prediction interval. Since it is a 95 % prediction interval we say that we expect that 5 of 100 observations will be outside the interval it is not unresonable that 1 of 10 is outside the interval.

\newpage

# Assignment 2

```{r}
assignment2 <- function(data){
    old <- par(mfrow = c(2, 2))
    acf(data, lag.max = 40, main="Data ACF")
    pacf(data, lag.max = 40, main="Data PACF")
    acf(diff(data, lag = 1), lag.max = 40, main="Difference 1 Data ACF")
    pacf(diff(data, lag = 1), lag.max = 40, main="Difference 1 Data PACF")
    par(old)
}
```

## Chicken

```{r, echo=FALSE, fig.height=10}
assignment2(chicken)
```

### Data ACF
The ACF on the original data suggests an AR or ARMA model since the ACF tails off.

### Data PACF
The PACF on the original data cuts off after lag 1 suggesting an AR(1) model.

### Difference 1 Data ACF
After having performed difference of order 1 we can clearly see that there is a seasonal trend in the data. The ACF suggests a seasonality of 10 but it does not seem to tail off.

### Difference 1 Data PACF
The PACF indicates that ...

### Final Verdict
ARIMA(1, 0, 0) x $(\_, 1, \_)_10$

\newpage

## so2

```{r, echo=FALSE, fig.height=10}
assignment2(so2)
```

### Data ACF
The ACF tails off suggesting either an AR or ARMA model.

### Data PACF
The PACF tails off as well suggesting an ARMA model.

### Difference 1 Data ACF
The ACF after difference cuts off after lag 1 suggesting a MA(1) model.

### Difference 1 Data PACF
The PACF after difference tails off further suggesting a MA(1) model.

### Final Verdict
ARIMA(0, 1, 1)

\newpage

## EQcount

```{r, echo=FALSE, fig.height=10}
assignment2(EQcount)
```

### Data ACF
The ACF tails off suggesting an AR or ARMA model.

### Data PACF
The PACF cuts off after lag 1 suggesting AR(1) model.

### Difference 1 Data ACF
The ACF after difference cuts off after lag 1 suggesting a MA(1) model.

### Difference 1 Data PACF


### Final Verdict


\newpage

## HCT

```{r, echo=FALSE, fig.height=10}
assignment2(HCT)
```

### Data ACF
The ACF tails off suggesting either an AR or ARMA model.

### Data PACF
The PACF cuts off after lag 7 suggesting an AR(7) model.

### Difference 1 Data ACF
The ACF suggests seasonality that tails off after lag 7 suggesting an seasonality of 7.

### Difference 1 Data PACF
The PACF cuts off after 6 lags suggesting an AR(6) seasonality model.

### Final Verdict
ARIMA(7, 0, 0) x $(6, 1, 0)_7$

\newpage

# Assignment 3

```{r}
plot_helper <- function(data, title) {
    old <- par(mfrow=c(4, 1))
    plot(data, main=title)
    acf(data, lag.max=40, main="")
    pacf(data, lag.max=40, main="")
    qqnorm(data, main="", las=1)
    qqline(data)
    par(old)
}

test_helper <- function(data) {
    print(Box.test(data, lag = 1, type = "Ljung-Box"))
    print(suppressWarnings(adf.test(data)))
    e <- eacf(data)
}

fit_plot <- function(model) {
    pred <- predict(model, n.ahead=20, se.fit=TRUE)
    upper_band <- pred$pred + 1.96 * pred$se
    lower_band <- pred$pred - 1.96 * pred$se

    plot(c(model$x, pred$pred), type="l", xlim=c(500, length(oil) + 20), ylim=c(min(lower_band), max(upper_band)))
    lines(length(oil) + 1:20, upper_band, lty=2, col="red")
    lines(length(oil) + 1:20, lower_band, lty=2, col="red")
}
```

## a)

```{r}
loil <- log(oil)
doil <- diff(oil)
ddoil <- diff(oil, 2)
dloil <- diff(loil)
ddloil <- diff(loil, 2)
```

\newpage

```{r, echo=FALSE, fig.height=10}
plot_helper(oil, "Oil")
```

```{r, echo=FALSE, fig.height=10}
plot_helper(doil, "1 Difference Oil")
```

```{r, echo=FALSE, fig.height=10}
plot_helper(ddoil, "2 Difference Oil")
```

```{r, echo=FALSE, fig.height=10}
plot_helper(dloil, "1 Difference log Oil")
```

```{r, echo=FALSE, fig.height=10}
plot_helper(ddloil, "2 Difference log Oil")
```

\newpage

```{r}
test_helper(doil)
test_helper(ddoil)
test_helper(dloil)
test_helper(ddloil)
```

\newpage

```{r}
fit1 <- Arima(loil, order=c(0, 2, 1))
fit1

fit_plot(fit1)
```

```{r}
fit2 <- Arima(loil, order=c(0, 1, 3))
fit2

fit_plot(fit2)
```

\newpage

```{r, fig.height=10, echo=FALSE}
old <- par(mfrow=c(4, 1))
plot(oil, main="Oil")
plot(diff(oil), main="Difference 1 Oil")
plot(dloil, main="Difference 1 Log Oil")
plot(ddloil, main="Difference 2 Log Oil")
par(old)
```

Clearly difference log is the data we should work with. bla, bla, ...


```{r, fig.height=10, echo=FALSE}
old <- par(mfrow=c(4, 1))
acf(dloil, lag.max=40, main="Difference 1 Log Oil ACF")
pacf(dloil, lag.max=40, main="Difference 1 Log Oil PACF")
acf(ddloil, lag.max=40, main="Difference 2 Log Oil ACF")
pacf(ddloil, lag.max=40, main="Difference 2 Log Oil PACF")
par(old)
```

```{r}
eacf(dloil)
eacf(ddloil)
```

```{r, fig.height=10, echo=FALSE}
old <- par(mfrow=c(2, 1))
qqnorm(dloil, main="Difference 1 Log Oil", las=1)
qqline(dloil, main="")
qqnorm(ddloil, main="Difference 2 Log Oil", las=1)
qqline(ddloil, main="")
par(old)
```

\newpage

```{r}
fit1 <- Arima(loil, order=c(1, 1, 1))
fit1
fit2 <- Arima(loil, order=c(0, 1, 3))
fit2
fit3 <- Arima(loil, order=c(0, 2, 1))
fit3
```

\newpage

```{r}
complex_dist <- function(x) {
    sqrt(Re(x)^2 + Im(x)^2)
}

sapply(polyroot(c(1, -2, 1)), complex_dist)
sapply(polyroot(c(1, -1)), complex_dist)
```

```{r}
fit_plot(fit3)
```

\newpage

## b)

```{r}
lunemp <- log(unemp)
dunemp <- diff(unemp)
ddunemp <- diff(dunemp, 2)
dlunemp <- diff(lunemp)
ddlunemp <- diff(lunemp, 2)
```

```{r, fig.height=10, echo=FALSE}
old <- par(mfrow=c(4, 1))
plot(unemp, main="Unemp")
plot(dunemp, main="Difference 1 Unemp")
plot(dlunemp, main="Difference 1 Log Unemp")
plot(ddlunemp, main="Difference 2 Log Unemp")
par(old)
```

Clearly difference log is the data we should work with. bla, bla, ...


```{r, fig.height=10, echo=FALSE}
old <- par(mfrow=c(4, 1))
acf(dlunemp, lag.max=40, main="Difference 1 Log Unemp ACF")
pacf(dlunemp, lag.max=40, main="Difference 1 Log Unemp PACF")
acf(ddlunemp, lag.max=40, main="Difference 2 Log Unemp ACF")
pacf(ddlunemp, lag.max=40, main="Difference 2 Log Unemp PACF")
par(old)
```

```{r}
eacf(dlunemp)
eacf(ddlunemp)
```

```{r, fig.height=10, echo=FALSE}
old <- par(mfrow=c(2, 1))
qqnorm(dlunemp, main="Difference 1 Log Unemp", las=1)
qqline(dlunemp, main="")
qqnorm(ddlunemp, main="Difference 2 Log Unemp", las=1)
qqline(ddlunemp, main="")
par(old)
```

\newpage

```{r}
fit1 <- Arima(lunemp, order=c(1, 1, 1))
fit1
fit2 <- Arima(lunemp, order=c(0, 1, 3))
fit2
fit3 <- Arima(lunemp, order=c(0, 2, 1))
fit3
```
